---
title: "Douglas-complex fire smoke and cardiopulmonary morbidity"
author: "Ryan Gan"
date: "11/3/2017"
output: html_document
---

##1. Introduction

In the summer of 2013, the Douglas-Complex fires occured in southwest Oregon. Oregonians in this part of the state were at risk of exposure to extreme levels of particulate matter (PM). This project aim is to determine if there is an association to smoke from the Douglas-Complex fires and acute cardiopulmonary morbidity in the state of Oregon.

Data came from the Oregon All Payer All Claims Database (APAC) in the year 2013. The APAC records the health care billing data for Oregon's insured populations. APAC include individual billing records for both diagnoses codes (International Classification of Diseases, Clinical Modification (ICD-9-CM) diagnoses codes) and pharmacy codes (National Drug Codes (NDC)). 

Our previous research that found an 
association with wildfire smoke and respiratory outcomes in Washington state in 2012 wildfire season using a novel estimate of smoke concentration, geographically weighted ridge regression (GWR) guided the methodological approaches used in this project. As Oregon contains pharmacy records, we evaluate the association between smoke and respiratory rescue medications (beta 2 agonists) (abbreviate to SABA). 

*Research question*
We evaluated the association between smoke concentrations using the GWR method and cardiopulmonary morbidity, including ED/urgent care visits and SABA fills in Oregon state during the 2013 wildfire season.

This markdown document contains the code and results that were used to address this research question.

Packages used: tidyverse, survival.
```{r setup, echo =F, warning=F, message=F}
library(tidyverse) # general data wrangle
library(survival) # conditional logistic regression
library(geofacet)

# knitr options
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```


##2. Wildfire smoke descriptive characteristics

Smoke exposure is represented by ZIP code-level population-weighted PM~2.5~ estimated using geographically-weighted ridge regression (GWR), where we have subtracted off the background median PM~2.5~ for the month where no smoke was present overhead in the atmospheric column using HMS. It was estimated every day for the wildfire season of May to October, 2013.

We've decided to represent descriptive characteristics of smoke by county-level as we can represent the location and temporal variation of smoke exposure better than we could at the ZIP code level.

```{r smoke data import and wrangle}
# pm path
pm_path <- paste0("../../data/pm/2013-oregon_county_pm25.csv")
# county pm
county_pm_df <- read_csv(pm_path) %>% 
  mutate(county = stringr::str_replace(county, "[.]", ""))

# fire locations
fire_locations <- read_csv("../../data/pm/2013-oregon_firelocations.csv") %>% 
  mutate(area = sqrt(total_acres/1000))

# estimate county smoke 
county_smk_count <- county_pm_df %>% 
  # binary smoke variable at >10 ug/m^3
  mutate(gwr_smk10 = ifelse(geo_smk_pm >= 10, 1, 0)) %>% 
  group_by(county) %>% 
  summarise(smk_day10 = sum(gwr_smk10)) %>% 
  # lower case county name and remove "." to match spatial df
  mutate(subregion = tolower(stringr::str_replace(county, "[.]", " ")))

# extract oregon map data
or_county_spatial_df <- map_data("county", "oregon") %>% 
  right_join(county_smk_count, by = "subregion")

# use the map function from the maps package to get the polygon data
county_poly <- maps::map("county", "oregon", plot=FALSE, fill = TRUE)
# find centroids
county_centroids <- maps:::apply.polygon(county_poly, maps:::centroid.polygon)

# create a data frame for graphing out of the centroids of each polygon
county_centroids <- county_centroids[!is.na(names(county_centroids))]
centroid_array <- Reduce(rbind, county_centroids)

county_text <- reduce(county_centroids, rbind) %>% 
  as_data_frame() %>% 
  rename(long = V1, lat = V2) %>% 
  mutate(county = stringr::str_sub(
    stringr::str_to_title(names(county_centroids)), start=8L)) 

# import msa location file from
# https://www.statcrunch.com/app/index.php?dataid=1232319
msa_path <- paste0("../instructions/msa_coordinate.csv")
msa_df <- read_csv(msa_path) 

msa_location <- msa_df %>%
  filter(grepl('OR', msa_df$MSA)) %>%
  filter(MSA != "Albany, OR" &
         MSA != "Grants Pass, OR")

# create msa label data frame
msa_text <- msa_location %>%
  select(MSA, Longitude, Latitude) 
  
msa_text$MSA <- gsub(",.*", "", msa_text$MSA)
```

We summarize smoke exposure two ways: as a count of days where smoke PM~2.5~ is > 10 ug/m^3 (to increase the specificity to identify areas impacted by smoke) at the county level, and time-series of county-population weighted PM~2.5~ over May to October 2013. 

### Number of days in Oregon where smoke PM~2.5~ exceeded 10 ug/m^3. 

```{r smoke days map}

smoke_map <- ggplot(or_county_spatial_df, aes(x=long,y=lat, group=group)) +
  
  # fill with number of smoke days
  geom_polygon(aes(fill =smk_day10)) +
  scale_fill_gradient2(expression("Smoke Days > 10 µg/m"^3),
    low = "white", mid= "#00dbde", high="#6a3093", midpoint = 15,
    guide = guide_colorbar(direction="horizontal", title.position = "top",
                           title.hjust = 0.5, barwidth = 20)) +
  # add county path on top
  geom_path(colour="#4ac29a") +
  # adding fire locations in
  geom_point(data=fire_locations, 
    aes(x=lon, y=lat, size = area,
    shape = "Fire Locations"), shape = 17, color = "#ff0084", alpha = 0.7,
    show.legend = FALSE, inherit.aes = FALSE) +
  # add county text 
  geom_text(data = county_text, 
    aes(x=long, y=lat, label = county, group="county"), 
    colour = "#33001b", size = 2.5) +
  # add metropolitan
  geom_point(data=msa_location, 
    aes(x=Longitude, y=Latitude, size = 1.5,
    shape = "MSA Locations"), shape = 19, color = "black", alpha = 0.7,
    show.legend = FALSE, inherit.aes = FALSE) +
  # add msa text
  geom_text(data = msa_text, 
    aes(x=Longitude + c(0.15,0.1,0.1,-0.05,0.2,0.1), y=Latitude + c(0.2,0.1,0.1,-0.1,0.1,0.1), label = MSA, group="MSA"), 
    colour = "#33001b", size = 2.5) +
  scale_color_manual(values = c("County" = '#ff0084','MSA' = '#33001b')) + 
  scale_shape_manual(values = c('County' = 17, 'MSA' = 19)) +
  
  xlab("Longitude") +
  ylab("Latitude") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(colour=NA, fill=NA),
  legend.position = "bottom") 
 
  
  # scale_colour_manual(name="Error Bars",values=cols) + 
  # scale_fill_manual(name="Bar",values=cols)

# scale_fill_manual(name = "Legends",
#                     values = c("#ff0084", "#33001b"),
#                     labels = c("County", "MSA"),
#                     guide = F) +

smoke_map

write_csv(or_county_spatial_df, "./or_county_spatial_df.csv")
# save figure
#ggsave(filename = "./smoke_map.png", plot = smoke_map)
```

Most of the heavy smoke days were in the south-west part of Oregon. Triangles are fires; had them in the legend, but the legend looked weird so I took them out.

